---
title: "day1"
author: "Amelia Ritger"
date: "4/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messages = FALSE)
```

Geospatial data in R workshop, Day 1

https://ucsbcarpentry.github.io/2022-04-07-ucsb-r-geospatial/01-raster-structure/

```{r}
library(raster)
library(rgdal)
library(tidyverse) #ggplot and dplyr
```

## PART 1 Intro to rasters

Get geospatial data for specific image
```{r}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

#open first raster
dsm_harv <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")
dsm_harv

summary(dsm_harv)
summary(dsm_harv, maxsamp=ncell(dsm_harv))

#turn dsm_harv into tidy dataframe with an xy coord system
dsm_harv_df <- as.data.frame(dsm_harv, xy=TRUE)
str(dsm_harv_df)
```

Plot the first raster (raster = continuous dataset and each pixel has a value from 0 to 255; vector = lines or points and polygons and noncontinuous dataset)
```{r}
ggplot() +
  geom_raster(data=dsm_harv_df, aes(x=x, y=y, fill=HARV_dsmCrop)) + #get fill= from str(dsm_harv_df), tells you value of elevation associated with each pixel
  scale_fill_viridis_c() +
  coord_quickmap()

#quickplot
plot(dsm_harv)
```

CRS formatting
```{r}
#view CRS string (aka the projection)
crs(dsm_harv)
resm(dsm_harv) #see resolution and pixel area

#find min and max values of dataset
minvalue(dsm_harv)
maxvalue(dsm_harv)

#set the min and max values (tell R to put them in order)
dsm_harv <- setMinMax(dsm_harv)
```

Raster bands (single or multi - see workshop website)
```{r}
nlayers(dsm_harv) #aka this is a single band
```

See if values fall where they should be
```{r}
ggplot(data = dsm_harv_df, aes(HARV_dsmCrop)) +
  geom_histogram(bins=40)
```

Challenge
```{r}
GDALinfo("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

#Does this file have the same CRS as dsm_harv?

#YES

#what is the NoDataValue?

#-9999

#What is the resolution of the raster data?

# 1x1? 

#How large would a 5x5 pixel area be on the Earth's surface?

#units in meters, so 5x5 m

#Is the file a multi or single raster band?

#single
```

## PART 2 Plot raster data

```{r}
#Create a new variable within df
dsm_harv_df <- dsm_harv_df %>%
  mutate(fct_elevation = cut(HARV_dsmCrop, breaks=3)) #make three even breaks of the elevation data

#identify break points
unique(dsm_harv_df$fct_elevation)

summary(dsm_harv_df)

ggplot(data=dsm_harv_df, aes(fct_elevation)) +
  geom_bar()

#identify break points
unique(dsm_harv_df$fct_elevation)
```

Create custom bins
```{r}
#set bin values 
custom_bins <- c(300, 350, 400, 450)

dsm_harv_df <- dsm_harv_df %>%
  mutate(fct_elevation2 = cut(HARV_dsmCrop, breaks=custom_bins)) #make three even breaks of the elevation data

summary(dsm_harv_df)
```
Create raster with terrain colors identified
```{r}
ggplot(data=dsm_harv_df, aes(x=x, y=y, fill=fct_elevation2)) +
  geom_raster() +
  scale_fill_manual(values=terrain.colors(3)) + 
  coord_quickmap()

my_colors <- terrain.colors(3)

ggplot(data=dsm_harv_df, aes(x=x, y=y, fill=fct_elevation2)) +
  geom_raster() +
  scale_fill_manual(values=my_colors, name = "Elevation") + 
  theme(axis.title = element_blank())
  coord_quickmap()
```

Layering rasters
```{r}
dsm_hill_harv <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif")
dsm_hill_harv

dsm_hill_harv_df <- as.data.frame(dsm_hill_harv, xy=TRUE)
str(dsm_hill_harv_df)
```

Plot the hillshape 
```{r}
ggplot(data=dsm_hill_harv_df, aes(x=x, y=y, alpha=HARV_DSMhill)) +
  geom_raster() +
  scale_alpha(range=c(0.15, 0.65), guide="none") +
  coord_quickmap()
```

Combine hill data and elevation data in one plot
```{r}
ggplot() +
  geom_raster(data=dsm_harv_df, aes(x=x, y=y, fill=HARV_dsmCrop)) +
  geom_raster(data=dsm_hill_harv_df, aes(x=x, y=y, alpha=HARV_DSMhill)) +
  scale_fill_viridis_c() +
  scale_alpha(range=c(0.15, 0.65), guide="none") +
  ggtitle("Elevation with hillshade") +
  coord_quickmap()
```

## PART 3 Reproject raster data 

Using DTM data (ditigal terrain model, bare earth without trees or houses etc), not DSM data (digital surface model, LIDAR source with bumps like trees but not actual ground)
```{r}
dtm_harv <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DTM/HARV_dtmCrop.tif")
dtm_hill_harv <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DTM/HARV_DTMhill_WGS84.tif")

dtm_harv_df <- as.data.frame(dtm_harv, xy=TRUE)
dtm_hill_harv_df <- as.data.frame(dtm_hill_harv, xy=TRUE)

crs(dtm_harv)
crs(dtm_hill_harv)

compareCRS(dtm_harv, dtm_hill_harv)
```

Rasters have to be the same CRS to overlay or layer!!

Reproject the raster to make the data match
```{r}
dtm_hill_reprojected <- projectRaster(dtm_hill_harv, crs=crs(dtm_harv))

#make a new df
dtm_hill_reprojected_df <- as.data.frame(dtm_hill_reprojected, xy=TRUE)

str(dtm_hill_reprojected_df) #check new projected df structure

ggplot() +
     geom_raster(data = dtm_harv_df , 
                 aes(x = x, y = y, 
                  fill = HARV_dtmCrop)) + 
     geom_raster(data = dtm_hill_reprojected_df, 
                 aes(x = x, y = y, 
                   alpha = HARV_DTMhill_WGS84)) +
     scale_fill_gradientn(name = "Elevation", colors = terrain.colors(10)) + 
     coord_quickmap()

#y axis is the number of meters away from the origin of the projection
```

## PART 4 Raster calculations

DSM - DTM = CHM (canopy height model)

let's start to build it!
```{r}
dsm_harv_df
str(dsm_harv_df)

#plot terrain model
ggplot() +
  geom_raster(data=dtm_harv_df, aes(x=x, y=y, fill=HARV_dtmCrop)) +
  scale_fill_gradientn(colors=terrain.colors(10)) +
  coord_quickmap()

#plot surface model
ggplot() +
  geom_raster(data=dsm_harv_df, aes(x=x, y=y, fill=HARV_dsmCrop)) +
  scale_fill_gradientn(colors=terrain.colors(10)) +
  coord_quickmap()

#we didn't have enough time to do CHM but look at website to finish it up...
```

Multi band color
```{r}
rgb_band1_harv <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/RGB_Imagery/HARV_RGB_Ortho.tif")

nbands(rgb_band1_harv)

rgb_band1_harv_df <- as.data.frame(rgb_band1_harv, xy=TRUE)

ggplot() +
  geom_raster(data=rgb_band1_harv_df, aes(x=x, y=y, alpha=HARV_RGB_Ortho)) +
  coord_quickmap()

#new function for color
rgb_stack_harv <- stack("data/NEON-DS-Airborne-Remote-Sensing/HARV/RGB_Imagery/HARV_RGB_Ortho.tif")
rgb_stack_harv #check it out, there are three bands AKA RGB!
```

